---
title: "Digital Divide in the US"
author: "Mustafa - Liam"
format: html
---

## Introduction

It goes without saying that the internet has grown at an exponential rate, providing more
access to information and opportunities than ever before. From the stock market to online SAT
resources on the internet has become the medium of the instruments of civilization. However,
what about the people who do not have equal access to the Internet? We looked at data from
across the United States, to see how access to broadband Internet impacted factors such as
education attainment, and financial prosperity. There is little digestible information on broadband
access and educational or economic outcomes, so the main point of this project is to combine
different data sets, so that we can take a closer look at various economic, educational and
geographic factors that impact access to vital Internet resources. It is very likely that Internet
access substantially impacts life outcomes, but it is good practice to look at the evidence before
making a conclusion. We will go through representations of data, but first we should go over the
origins of the data.

---
title: "Digital Divide in the US"
author: "Mustafa - Liam"
format: html

library(tidyverse)
library(janitor)
library(sf)
library(tmap)
library(geodata)
library(tinytable)
library(tigris)
library(dplyr)
library(stringr)
library(plotrix)
library(ggplot2)
library(reshape2)
---

## Introduction

It goes without saying that the internet has grown at an exponential rate, providing more
access to information and opportunities than ever before. From the stock market to online SAT
resources on the internet has become the medium of the instruments of civilization. However,
what about the people who do not have equal access to the Internet? We looked at data from
across the United States, to see how access to broadband Internet impacted factors such as
education attainment, and financial prosperity. There is little digestible information on broadband
access and educational or economic outcomes, so the main point of this project is to combine
different data sets, so that we can take a closer look at various economic, educational and
geographic factors that impact access to vital Internet resources. It is very likely that Internet
access substantially impacts life outcomes, but it is good practice to look at the evidence before
making a conclusion. We will go through representations of data, but first we should go over the
origins of the data.


```{r}
## Reading and cleaning the Data

broadband <- read_csv("https://raw.githubusercontent.com/Mustafa-Ali-code/stat100project/refs/heads/main/datasets/broadband_data_2020October.csv")
education <- read_csv("https://raw.githubusercontent.com/Mustafa-Ali-code/stat100project/refs/heads/main/datasets/Education2023.csv")
population <- read_csv("https://raw.githubusercontent.com/Mustafa-Ali-code/stat100project/refs/heads/main/datasets/PopulationEstimates.csv")
poverty <- read_csv("https://raw.githubusercontent.com/Mustafa-Ali-code/stat100project/refs/heads/main/datasets/Poverty2023.csv")
unemployment <- read_csv("https://raw.githubusercontent.com/Mustafa-Ali-code/stat100project/refs/heads/main/datasets/Unemployment2023.csv")

broadband <- clean_names(broadband)
education <- clean_names(education)
population <- clean_names(population)
poverty <- clean_names(poverty)
unemployment <- clean_names(unemployment)


broadband_clean <- broadband |>
  select(fips = county_id, state = st, county_name,
         broadband_availability_per_fcc, broadband_usage)

education_clean <- education |>
  filter(attribute == "Percent of adults with a bachelor's degree or higher, 2019-23") |>
  select(fips = fips_code, pct_bachelors_2023 = value)

population_clean <- population |>
  filter(attribute == "POP_ESTIMATE_2023") |>
  select(fips = fip_stxt, population_2023 = value)

poverty_clean <- poverty |>
  filter(attribute %in% c("PCTPOVALL_2023", "MEDHHINC_2023")) |>
  pivot_wider(names_from = attribute, values_from = value) |>
  rename(fips = fips_code,
         poverty_rate_2023 = PCTPOVALL_2023,
         median_income_2023 = MEDHHINC_2023)

unemployment_clean <- unemployment |>
  filter(attribute == "Unemployment_rate_2023") |>
  select(fips = fips_code, unemployment_rate_2023 = value)

```

```{r}
## Merge all cleaned data on fips
final_data <- broadband_clean |>
  left_join(education_clean, by = "fips") |>
  left_join(population_clean, by = "fips") |>
  left_join(poverty_clean, by = "fips") |>
  left_join(unemployment_clean, by = "fips")

final_data <- final_data |>
  mutate(
    broadband_availability_per_fcc = as.numeric(broadband_availability_per_fcc),
    broadband_usage = as.numeric(broadband_usage)
  )
```

```{r}
## Graph 1: Broadband Usage vs. Broadband Availability
ggplot(final_data, aes(x = broadband_availability_per_fcc, y = broadband_usage, na.rm = TRUE)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Broadband Usage vs. Availability (per FCC)",
    x = "Broadband Availability (%)",
    y = "Broadband Usage (%)"
  ) +
  theme_minimal()


counties_sf <- st_read("cb_2023_us_county_5m.shp") |>
  clean_names()

counties_sf <- counties_sf |>
  mutate(fips = geoid)

final_data <- final_data |>
  mutate(fips = str_pad(fips, width = 5, pad = "0"))

map_data <- left_join(counties_sf, final_data, by = "fips")

map_data <- map_data |> 
  filter(!str_sub(fips, 1, 2) %in% c("02", "15", "72"))

tmap_mode("view")

tm_shape(map_data) +
  tm_polygons("broadband_usage",
              fill.scale = tm_scale_intervals(style = "jenks"),
              fill.legend = tm_legend(title = "Broadband Usage (%)"),
              col_alpha = 0.1) +
  tm_title("Broadband Usage by County") +
  tm_crs("auto")
```

```{r}
# Graph 2: Usage vs. Median Income
income_pie_data <- final_data |>
  mutate(income_group = case_when(
    median_income_2023 < 50000 ~ "<50k",
    median_income_2023 >= 50000 & median_income_2023 < 60000 ~ "50–60k",
    median_income_2023 >= 60000 & median_income_2023 < 70000 ~ "60–70k",
    median_income_2023 >= 70000 & median_income_2023 < 80000 ~ "70–80k",
    median_income_2023 >= 80000 ~ "80k+",
    TRUE ~ NA_character_
  )) |>
  group_by(income_group) |>
  summarize(total_usage = sum(broadband_usage, na.rm = TRUE)) |>
  na.omit() |>
  arrange(income_group)

slices <- income_pie_data$total_usage
percentages <- round(slices / sum(slices) * 100, 1)
labels <- paste0(income_pie_data$income_group, " (", percentages, "%)")

pie3D(
  slices,
  labels = NA,
  explode = 0.1,
  main = "Broadband Usage by Income Group (3D Pie Chart)",
  radius = 1.4,
  col = rainbow(length(slices)),
  labelcex = 1
)

legend("topright", legend = labels, fill = rainbow(length(slices)), cex = 1.1, bty = "n")
```

```{r}
## Graph 3: Usage vs. Education (% Bachelor's Degree or Higher)
final_data |>
  filter(!is.na(pct_bachelors_2023)) |>  # First, drop pre-existing NAs
  mutate(edu_group = cut(
    pct_bachelors_2023,
    breaks = c(0, 10, 20, 30, 40, 50, Inf),
    labels = c("<10%", "10–20%", "20–30%", "30–40%", "40–50%", "50%+"),
    right = FALSE  # Optional: adjust if bins should be inclusive on the left instead
  )) |>
  filter(!is.na(edu_group)) |>  # Drop any NAs created by cut()
  group_by(edu_group) |>
  summarize(avg_usage = mean(broadband_usage, na.rm = TRUE)) |>
  ggplot(aes(x = edu_group, y = avg_usage)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Broadband Usage by Education Level",
    x = "% with Bachelor's Degree or Higher (Grouped)",
    y = "Average Broadband Usage (%)"
  ) +
  theme_minimal()
```

```{r}
## Table: Top 10 Counties by Usage
top10_table <- final_data |>
  select(county_name, state, broadband_usage, median_income_2023,
         pct_bachelors_2023, poverty_rate_2023) |>
  mutate(broadband_usage = round(broadband_usage * 100, 1)) |>
  arrange(desc(broadband_usage)) |>
  slice_head(n = 10) |>
  rename(
    `County` = county_name,
    `State` = state,
    `Usage (%)` = broadband_usage,
    `Median Income` = median_income_2023,
    `% Bachelor's Degree` = pct_bachelors_2023,
    `Poverty Rate (%)` = poverty_rate_2023
  )

tt(top10_table)
```

```{r}
# Table of summary stats
summary_full <- final_data |>
  summarize(
    `Broadband Usage` = list(quantile(broadband_usage, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)),
    `Availability (FCC)` = list(quantile(broadband_availability_per_fcc, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)),
    `Median Income` = list(quantile(median_income_2023, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)),
    `% with Bachelor's` = list(quantile(pct_bachelors_2023, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)),
    `Poverty Rate` = list(quantile(poverty_rate_2023, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)),
    `Unemployment Rate` = list(quantile(unemployment_rate_2023, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE))
  ) |>
  tidyr::pivot_longer(cols = everything(), names_to = "Variable", values_to = "Values") |>
  mutate(
    Min = sapply(Values, function(x) x[1]),
    Q1 = sapply(Values, function(x) x[2]),
    Median = sapply(Values, function(x) x[3]),
    Q3 = sapply(Values, function(x) x[4]),
    Max = sapply(Values, function(x) x[5])
  ) |>
  select(Variable, Min, Q1, Median, Q3, Max)

tt(summary_full)
```

```{r}
# Correlation Heatmap
correlation_data <- final_data |>
  select(broadband_usage, median_income_2023, pct_bachelors_2023, poverty_rate_2023, unemployment_rate_2023)

corr_matrix <- cor(correlation_data, use = "complete.obs")

corr_matrix_melted <- melt(corr_matrix)

ggplot(corr_matrix_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1,1)) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 5) +  # Add correlation values on the tiles
  theme_minimal() +
  labs(
    title = "Correlation Heatmap for Broadband Usage Factors",
    x = "Variables",
    y = "Variables",
    fill = "Correlation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#Broadband Usage by Region (North, South, East, West)
final_data <- final_data |>
  mutate(region = case_when(
    state %in% c("ME", "VT", "NH", "MA", "RI", "CT", "NY", "NJ", "PA", "DE", "MD", "DC", "VA") ~ "Northeast",
    state %in% c("OH", "IN", "IL", "WI", "MI", "IA", "MO", "KY", "MN", "ND", "SD", "NE") ~ "Midwest",
    state %in% c("WA", "OR", "CA", "ID", "MT", "WY", "CO", "NM", "AZ", "UT", "NV") ~ "West",
    state %in% c("NC", "SC", "GA", "FL", "AL", "MS", "TN", "AR", "LA", "OK", "TX") ~ "South",
    TRUE ~ "Other"
  ))

region_usage <- final_data |>
  group_by(region) |>
  summarize(average_usage = mean(broadband_usage, na.rm = TRUE))

ggplot(region_usage, aes(x = region, y = average_usage, fill = region)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Broadband Usage by Region", x = "Region", y = "Average Broadband Usage (%)") +
  theme_minimal()
```

```{r}
# Map with Usage vs. Availability (The difference)
map_data <- map_data |>
  mutate(diff_usage_availability = broadband_availability_per_fcc - broadband_usage)

tmap_mode("view")

tm_shape(map_data) +
  tm_borders() +
  tm_polygons("diff_usage_availability", 
              title = "Difference Between Availability and Usage (%)", 
              palette = "-RdBu",  # Using a diverging palette to show difference
              alpha = 0.6) +
  tm_layout(main.title = "Difference Between Broadband Availability and Usage")
```







